Simple and robust inference of connectomes using partial correlation coefficients

### 2014-06-10 article.tex
Merge both part on partial correlation

\textcolor{red}{Motivation pour l'inverse : retirer les liens indirects.
Pareil: proposition de AS  et certains choses pourraient être incorporées
dans la version de AJ.}

@@

Let us consider the neurons as random variables $\mathcal{X}_i$ which follows
a probability density function $p_{\mathcal{X}_i}$ and let us denote by
$p_{\mathcal{X}_i,\mathcal{X}_j|\mathcal{Z}}$ the joint probability
distribution between neurons $i$ and $j$ conditioned on a subset of neurons
$\mathcal{Z}$ (eventually empty). Let us define $\mathcal{V}$ the set of all
neurons $\mathcal{X}_1, \dots, \mathcal{X}_p$, and by $\mathcal{V}^{i,j}$ the
same set without neurons $\mathcal{X}_i$ and $\mathcal{X}_j$.

% Without considering indirect/direct --> Correlation matrix

First, if we want to highlight relationships between neurons without
considering the distinction between direct and indirect links, the Pearson
correlation matrix, i.e. each element $(i,j)$ of a such matrix is the
correlation coefficient between neurons $\mathcal{X}_i$ and $\mathcal{X}_j$, can computed.

% However, we only want direct links --> partial correlation

However the connectivity network to infer is only made of direct relationships
which can be defined as a relation between neurons that still exists when
conditioned on all other neurons or, in other words, there is not intermediary
between those two neurons.  Under the hypothesis that neurons follow a jointly
Gaussian distribution $p_{\mathcal{X}_1, \dots, \mathcal{X}_p} \sim
\mathcal{N}(\mu; \Sigma)$, two neurons $\mathcal{X}_i$ and $\mathcal{X}_j$ are
independent $\mathcal{X}_i\perp \mathcal{X}_j$ if the Pearson correlation
coefficient $\rho_{X_iX_j} = \frac{\Sigma_{ij}}{\sqrt{\Sigma_{ii}
\Sigma_{jj}}} = 0,$ and are conditionally independant to all others neurons
$\mathcal{X}_i \perp \mathcal{X}_j |\mathcal{V}^{i,j}$ if the partial
correlation coefficient (of order $p-2$) $\rho_{X_i, X_j | \mathcal{V}^{i,j}}
= \frac{\Sigma^{-1}_{ij}}{\sqrt{\Sigma^{-1}_{ii} \Sigma^{-1}_{jj}}} = 0.$
Therefore, by definition, the partial correlation coefficient between two
neurons only estimates the direct relationship between them.

@@

\vspace{1em}
\hrule
\vspace{1em}

% introduce the notion of dependence / independence
From a probabilistic point of view, the inference of the undirected network
can be formulated as inferring conditional dependences and
independences.  Even if \textit{correlation
does no imply causation}[todo cite], it gives insight on the underlying connectivity
network. For instance, the graph $A \rightarrow B \rightarrow C$ translates the
following dependances and independances: $A \not\perp B$, $A \not\perp B | C$,
$B \not\perp C$, $B \not\perp C | A$, but $A \perp C | B$ and  $A \not\perp C$.

Let us consider that each neuron signal $i$ is a random continuous variable
$X_i$ which follows a probability distribution $p_{X_i}$ and let us denote by
$p_{X_i,X_j|\mathcal{Z}}$ the  joint probability distribution of neurons $X_i$
and $X_j$ conditioned on a subset of variables $\mathcal{Z}$ (eventually
empty). Let us define $\mathcal{V}$ the set of all neurons $X_1, ..., X_p$ and
$\mathcal{V}^{-i,j}$ the same set without neurons $X_i$ and $X_j$.

Each neuron signal $i$ is a random continuous variable $X_i$, which follows a
probability density function $p_\mathcal{X_i}(x_i)$. Two neurons $X_i$, $X_j$
will be said conditionally independent to a set of neurons
$\mathcal{V}^{-i,j}$, denoted by  $X_i \perp X_j | \mathcal{V}^{-i,j}$,  if
$p_{X_i,X_j|\mathcal{V}^{-i,j}} = p_{X_i|\mathcal{V}^{-i,j}}
p_{X_j|\mathcal{V}^{-i,j}} $.

%for all real numbers $x_1,\ldots,x_p $ such that
%$p_{X_1,\ldots,X_p \backslash \left\{X_i, X_j\right\}}
%(x_1,\ldots,x_p \backslash \left\{x_i, x_j\right\}) > 0$. Note that this
%reciprocal is true only under additional hypotheses.

Under the hypothesis that the neurons follow a jointly Gaussian distribution
$p_{X_1,\ldots,X_p}(x_1,\ldots,x_p) \sim \mathcal{N}(\mu; \Sigma)$ of mean
vector $\mu$ and of covariance matrix $\Sigma$, two neurons $X_i$ and $X_j$
are independent $X_i \perp X_j$ if the Pearson correlation coefficient \[
\rho_{X_iX_j} = \frac{\Sigma_{ij}}{\sqrt{\Sigma_{ii} \Sigma_{jj}}} = 0, \]
and are conditionally independent to all others neurons $X_i \perp X_j |
X_1,\ldots,X_p \backslash \left\{X_i, X_j\right\}$ if the partial correlation
coefficient \[ \rho_{X_i, X_j | X_1,\ldots,X_p \backslash \left\{X_i,
X_j\right\}} = \frac{\Sigma^{-1}_{ij}}{\sqrt{\Sigma^{-1}_{ii}
\Sigma^{-1}_{jj}}} = 0. \]

% Correlation => Issue indirect effect
% Partial correlation => better estimate of the graph  => Symmetric measure
While the correlation coefficient is unable to distinguish between
direct and indirect interaction, the partial correlation coefficient
aims to infer only direct ones through the conditioning over all the other
variables. In previous study \cite{shipley2002cause}, it has been suggested to
consider correlation coefficients of all possible orders,
i.e. conditioned on every subsets of the set of $p-2$ other variables. This
would take into account multiple indirect paths between a given pair of
variables. An obvious limitation to estimate correctly high-order coefficients
is the number of samples. However given the high number of samples $n$
in the Connectomics challenge, we only consider the highest order of partial
correlation coefficient.

>>>>

@@

Let us consider the neurons as random variables $\mathcal{X}_i$ which follows
a probability density function $p_{\mathcal{X}_i}$ and let us denote by
$p_{\mathcal{X}_i,\mathcal{X}_j|\mathcal{Z}}$ the joint probability
distribution between neurons $i$ and $j$ conditioned on a subset of neurons
$\mathcal{Z}$ (possibly empty). Two neurons $X_i$, $X_j$
will be said conditionally independent to a set of neurons
$\mathcal{Z}$ (possibly empty), denoted by  $X_i \perp X_j | \mathcal{Z}$,  if
$p_{\mathcal{X}_i,\mathcal{X}_j|\mathcal{Z}} = p_{\mathcal{X}_i|\mathcal{Z}}
p_{\mathcal{X}_j|\mathcal{Z}}$. The reciprocal is not true without
additional hypotheses.  Let us define $\mathcal{V}$ the set of all
neurons $\mathcal{X}_1, \dots, \mathcal{X}_p$, and by $\mathcal{V}^{-i,j}$ the
same set without neurons $\mathcal{X}_i$ and $\mathcal{X}_j$.

The connectivity network to infer is only made of direct relationships
which defined relation between neurons that still exists when
conditioned on all other neurons. In other words, there is not intermediary
between those two neurons. Under the hypothesis that neurons follow a jointly
Gaussian distribution $p_{\mathcal{X}_1, \dots, \mathcal{X}_p} \sim
\mathcal{N}(\mu; \Sigma)$, two neurons $\mathcal{X}_i$ and $\mathcal{X}_j$ are
independent $\mathcal{X}_i\perp \mathcal{X}_j$ if the Pearson correlation
coefficient $\rho_{X_iX_j} = \frac{\Sigma_{ij}}{\sqrt{\Sigma_{ii}
\Sigma_{jj}}} = 0,$ and are conditionally independant to all others neurons
$\mathcal{X}_i \perp \mathcal{X}_j |\mathcal{V}^{i,j}$ if the partial
correlation coefficient (of order $p-2$) $\rho_{X_i, X_j | \mathcal{V}^{i,j}}
= \frac{\Sigma^{-1}_{ij}}{\sqrt{\Sigma^{-1}_{ii} \Sigma^{-1}_{jj}}} = 0$,
where $\Sigma^{-1}_$ is the inverse of the covariance matrix.
Therefore, by definition, the partial correlation coefficient between two
neurons only estimates the direct relationship between them.

@@
